\chapter*{Copyrights}

This document gives practical examples of the use of Elmer,
a finite element software for multiphysical problems.
The copyright if this document and Elmer software belong to
CSC -- IT Center for Science, Finland, 1995--2008. 

Elmer program is free software; you can redistribute it and/or
modify it under the terms of the GNU General Public License
as published by the Free Software Foundation; either version 2
of the License, or (at your option) any later version.

This program is distributed in the hope that it will be useful,
but WITHOUT ANY WARRANTY; without even the implied warranty of
MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
GNU General Public License for more details.

CSC assumes no responsibility or liability on any errors or inaccuracies in 
Elmer program or documentation. Any kind of material damages or other indirect
consequences resulting from any Elmer part, documentation discrepancies and 
errors or non-anticipated program behavior are limited to the value of 
appropriate products purchased from CSC. 

This document is for informational use only. All information and specifications
given have been carefully prepared by the best efforts of CSC, and are believed
to be true and accurate as of time publishing. CSC reserves the right to 
modify Elmer and its documents without notice. \\  \mbox{} \\

\textbf{Additional software product copyrights included in Elmer}

Copyright, and License:

UMFPACK Version 4.3, Jan. 16, 2004. Copyright (c) 2004 by Timothy
A. Davis, University of Florida, davis@cise.ufl.edu. All Rights
Reserved. 


UMFPACK License: 

Your use or distribution of UMFPACK or any derivative code implies
that you agree to this License. 

      THIS MATERIAL IS PROVIDED AS IS, WITH ABSOLUTELY NO WARRANTY
      EXPRESSED OR IMPLIED. ANY USE IS AT YOUR OWN RISK. 

      Permission is hereby granted to use or copy this program,
      provided that the Copyright, this License, and the Availability
      of the original version is retained on all copies. User
      documentation of any code that uses UMFPACK or any modified
      version of UMFPACK code must cite the Copyright, this License,
      the Availability note, and "Used by permission." Permission to
      modify the code and to distribute modified code is granted,
      provided the Copyright, this License, and the Availability note
      are retained, and a notice that the code was modified is
      included. This software was developed with support from the
      National Science Foundation, and is provided to you free of
      charge. 

